\section{Truth as Fixed Point: RG Flow as Logical Semantics}
\label{sec:truth-fixed-point}

Having established the interacting positive logic framework in Section~\ref{sec:formal-systems}, we now present the core innovation of this work: truth can be understood as a fixed point under renormalization group flow. This connection between renormalization and logic provides the foundation for our computational semantics. Everything in this section is domain-neutral; whenever we later speak about truth in the physics or learning domain we will simply reuse these fixed-point results with different interpretations of the observable space $\mathsf{Obs}$.

\subsection{Truth Fixed Points in the L/B/R Structure}

The interacting positive logic provides a natural framework for understanding truth as fixed points. The L/B/R structure creates boundaries where truth can be evaluated, with the bulk providing the computational dynamics.

\begin{theorem}[Truth as Fixed Point in L/B/R Structure]
\label{thm:truth-fixed-point-lbr}
Let $(\mathsf{Obs},\preceq)$ be an $\omega$-cpo with L/B/R structure. Assume $\mathcal{R}:\mathsf{Obs}\to\mathsf{Obs}$ is monotone and $\omega$-continuous with respect to the equality hierarchy $\equiv_L, \equiv_B, \equiv_R, \equiv_{\text{loc}}, \equiv_{\text{meta}}$. Then $\mathrm{lfp}(\mathcal{R})=\sup_n \mathcal{R}^n(\bot)$ exists (Kleene) and respects the L/B/R boundaries.
\end{theorem}

\begin{proof}[Proof Sketch]
The proof follows from Kleene's fixed point theorem, but now incorporates the L/B/R structure. Since $\mathcal{R}$ is monotone and $\omega$-continuous, the sequence $\{\mathcal{R}^n(\bot)\}_{n \geq 0}$ forms a chain in the $\omega$-cpo. The L/B/R boundaries are preserved under the RG flow because:
\begin{itemize}
\item Left boundary $L$: Truth evaluation via $\equiv_L$ equality
\item Bulk $B$: Computational dynamics via $\equiv_B$ equality  
\item Right boundary $R$: Truth evaluation via $\equiv_R$ equality
\end{itemize}
By Kleene's theorem, this chain has a least upper bound which is the least fixed point, and the L/B/R structure is preserved throughout the iteration.
\end{proof}

\begin{remark}[Connection to Implementation]
\label{rem:implementation-connection-lbr}
The RG operator iteration corresponds to iteration of the equality checking procedures in our implementation. The monotonicity condition is satisfied by the semiring operations in the M3/M2/M1 hierarchy, and the completeness follows from the $\omega$-cpo structure of our observable space. The L/B/R boundaries are implemented as:
\begin{itemize}
\item \texttt{M2\_pgc.rkt}: Left and right boundary DSLs
\item \texttt{M3\_rules.rkt}: Bulk computational dynamics
\item \texttt{M2\_cert.rkt}: Equality hierarchy implementation
\end{itemize}
\end{remark}

\subsection{Regularization as Deformation of L/B/R Structure}

Regularization deforms the L/B/R structure in a specific way, analogous to how temperature deforms crystal structures in condensed matter physics:

\begin{definition}[Deformation of L/B/R Structure]
Regularization induces a deformation of the L/B/R structure:
\begin{itemize}
\item Undeformed system: Standard L/B/R boundaries with classical equality
\item Deformed system: L/B/R structure with RG flow semantics and equality hierarchy
\item Deformation parameter: The scale parameter $\Lambda$ and presentation gauges $\equiv_{\text{scale}}, \equiv_{\text{hel}}$
\end{itemize}
\end{definition}

For our HEP-TH audience, this deformation should feel familiar: just as we deform field theories by introducing regulators (like dimensional regularization), we deform the L/B/R structure by introducing computational regulators. The key insight is that the same mathematical structure—deformation theory—applies to both contexts, with the L/B/R boundaries providing natural cut-off scales.

\subsection{Complete Equality Hierarchy (Six Layers)}

The logic features a sophisticated hierarchy of equality notions, aligned with the L/B/R structure. All observational equalities are computed modulo presentation gauges by default.

\begin{definition}[Presentation Gauges (Always Factored Out)]
\label{def:presentation-gauges}
The presentation gauges are always factored out internally:

Scale gauge $\equiv_{\text{scale}}$: least congruence with generators $t \equiv t \otimes_B (z \otimes_B \bar{z})$ for all closed bulk scalars $t$.

Helicity/phase gauge $\equiv_{\text{hel}}$: paired congruence generated by $(t,u) \sim (t \otimes_B z, u \otimes_B \bar{z})$ and $(t,u) \sim (t \otimes_B \bar{z}, u \otimes_B z)$.

All observational equalities below are computed modulo $\equiv_{\text{scale}}$ and $\equiv_{\text{hel}}$ by default. Let $\equiv_*$ denote their intersection.
\end{definition}

\begin{definition}[Five Observational Equalities (Complete)]
\label{def:observational-equalities}
For closed bulk scalars $t, u$, we define five notions of equality:

1. Left-local equality $\equiv_L$:
\[
t \equiv_L u \quad :\Leftrightarrow \quad \forall C_L[-]: B \to L \text{ built from } \{\iota_L, \oplus_L, \otimes_L\} \text{ and bulk ops: } C_L[t] \equiv C_L[u]
\]

2. Bulk equality $\equiv_B$:
\[
t \equiv_B u \quad :\Leftrightarrow \quad \forall C_B[-]: B \to B \text{ from } \{\oplus_B, \otimes_B, \text{ad}_i, F_{ij}, \mathsf{Gen4}\}: C_B[t] \equiv_* C_B[u]
\]

3. Right-local equality $\equiv_R$:
\[
t \equiv_R u \quad :\Leftrightarrow \quad \forall C_R[-]: B \to R \text{ built from } \{\iota_R, \oplus_R, \otimes_R\} \text{ and bulk ops: } C_R[t] \equiv C_R[u]
\]

4. Local-agreement equality $\equiv_{\text{loc}}$:
\[
t \equiv_{\text{loc}} u \quad :\Leftrightarrow \quad (t \equiv_L u) \land (t \equiv_R u)
\]

5. Metalogical global equality $\equiv_{\text{meta}}$:
\[
t \equiv_{\text{meta}} u \quad :\Leftrightarrow \quad \forall C[-]: B \to B \text{ built from entire signature: } C[t] \equiv_* C[u]
\]
\end{definition}

\begin{definition}[Bonus Layer: Reversible Equality $\equiv_\star$]
\label{def:reversible-equality}
The reversible equality $\equiv_\star$ is the largest congruence generated by invertible rewrites only:
\begin{itemize}
\item SMC structural isomorphisms (associativity/unit/symmetry)
\item Braided commutations $\text{ad}_i; \text{ad}_j \leftrightarrow \text{ad}_j; \text{ad}_i \otimes_B F_{ij}$ when $F_{ij}$ has an inverse
\item Elimination/insertions of $1_B$ and of paired $z/\bar{z}$ with zero net effect
\end{itemize}

Reversible Collapse Theorem: $\equiv_\star \subseteq (\equiv_{\text{loc}} \cap \equiv_{\text{meta}})$, and every $\equiv_\star$-step is computationally reversible (no information loss). In the flat subtheory ($F_{ij} = 1_B$), $\equiv_\star$ coincides with SMC isomorphism equivalence.
\end{definition}

\paragraph{Equality hierarchy completeness.} This six-layer equality hierarchy provides the complete specification for observational equalities in the L/B/R structure. The bonus layer $\equiv_\star$ captures the reversible, information-preserving core of the system, while the five observational equalities provide different perspectives on truth evaluation.

\subsection{Hierarchical Deformation via Complete Equality Hierarchy}

The complete six-layer equality hierarchy $\equiv_L, \equiv_B, \equiv_R, \equiv_{\text{loc}}, \equiv_{\text{meta}}, \equiv_\star$ provides natural understanding of how truth emerges through RG flow deformation. The deformation can be expressed as:
\[
\text{True}_\alpha = \lim_{\Lambda \to \infty} \mathcal{R}_\Lambda(\text{True}_{\alpha-1})
\]
where $\alpha$ ranges over ordinals, and each level respects the appropriate equality notion:
\begin{itemize}
\item $\text{True}_L$: Respects $\equiv_L$ equality (left boundary)
\item $\text{True}_B$: Respects $\equiv_B$ equality (bulk dynamics)
\item $\text{True}_R$: Respects $\equiv_R$ equality (right boundary)
\item $\text{True}_{\text{loc}}$: Respects $\equiv_{\text{loc}}$ equality (local agreement)
\item $\text{True}_{\text{meta}}$: Respects $\equiv_{\text{meta}}$ equality (metalogical global)
\item $\text{True}_\star$: Respects $\equiv_\star$ equality (reversible core)
\end{itemize}

\paragraph{Equality hierarchy in RG flow.} The six equality layers provide different perspectives on truth evaluation through RG flow, with the bonus layer $\equiv_\star$ capturing the reversible, information-preserving core of the system.

\begin{definition}[Directed Boundaries in L/B/R Structure]
\label{def:directed-boundaries-truth}
The L/B/R structure provides natural directed boundaries:
\begin{itemize}
\item Left boundary $L$: Input boundary with $\equiv_L$ equality
\item Bulk $B$: Computational dynamics with $\equiv_B$ equality
\item Right boundary $R$: Output boundary with $\equiv_R$ equality
\end{itemize}
The $\mathsf{G}_6$ primitive acts as a correlator between these boundaries, providing holographic structure.
\end{definition}

\begin{remark}[Holographic Renormalization in dS/CFT]
\label{rem:holographic-renorm-truth}
The L/B/R structure with $\mathsf{G}_6$ as correlator interprets as holographic renormalization in dS/CFT context (detailed in Section~\ref{sec:boundary-maps}). The bulk $B$ corresponds to the bulk spacetime, while $L$ and $R$ correspond to the boundary CFTs.
\end{remark}

\subsection{Spectral Gap and Computational Semantics}

\begin{definition}[Spectral Gap in Complete Equality Hierarchy]
\label{def:spectral-gap-truth}
The spectrum of the $\mathsf{Gen4}$ primitive has a natural symmetry-related gap between different equality layers:
\begin{itemize}
\item Kernel: Reversible computations (information preserving) - corresponds to $\equiv_\star$ equality
\item Co-kernel: Irreversible computations (information destroying) - corresponds to $\equiv_{\text{meta}} \setminus \equiv_\star$
\item Spectral Gap: The difference between reversible and irreversible computation, measured by the complete equality hierarchy
\end{itemize}
\end{definition}

This spectral gap is crucial for understanding the computational semantics of the $\mathsf{Gen4}$ primitive within the L/B/R structure and the complete equality hierarchy.

\subsection{Truth as RG Fixed Point}

\begin{theorem}[Truth as RG Fixed Point in Complete Equality Hierarchy]
\label{thm:truth-rg-fixed-point-lbr}
A statement $\phi$ is true if and only if it corresponds to a fixed point under RG flow within the L/B/R structure:
\[
\text{True}(\phi) \Leftrightarrow \lim_{\Lambda \to \infty} \mathsf{Gen4}(\llbracket\phi\rrbracket, \llbracket\bar{\phi}\rrbracket; \vec{q}, \Lambda) \text{ converges}
\]
where convergence of the RG flow corresponds to reversible computation (information preservation) and respects the complete equality hierarchy $\equiv_L, \equiv_B, \equiv_R, \equiv_{\text{loc}}, \equiv_{\text{meta}}, \equiv_\star$.
\end{theorem}

\subsection{Computational Semantics}

\begin{definition}[Computational Semantics in Complete Equality Hierarchy]
\label{def:computational-semantics-lbr}
The computational semantics of our framework assigns meaning to logical statements through RG flow behavior within the L/B/R structure and complete equality hierarchy:
\begin{itemize}
\item Truth: Converging RG flow (reversible computation) - corresponds to $\equiv_\star$ equality
\item Falsehood: Diverging RG flow (irreversible computation) - corresponds to $\equiv_{\text{meta}} \setminus \equiv_\star$
\item Undecidability: Marginal RG flow (undecidable computation) - corresponds to $\equiv_{\text{loc}} \setminus \equiv_\star$
\end{itemize}
Each semantic notion respects the appropriate equality layer in the complete six-layer hierarchy.
\end{definition}

\subsection{Summary and Outlook}

This section has established truth as a fixed point under RG flow within the L/B/R structure and complete equality hierarchy, providing the computational semantics for our framework. The key observations are:

\begin{enumerate}
\item Regularization deforms the L/B/R structure through RG flow
\item Complete six-layer equality hierarchy creates a hierarchy of truth predicates
\item L/B/R boundaries enable holographic renormalization with $\mathsf{Gen4}$ as correlator
\item The spectral gap distinguishes reversible from irreversible computation via complete equality hierarchy
\item Truth corresponds to converging RG flow (fixed points) respecting all six equality layers
\item Computational semantics assigns meaning through RG flow behavior within L/B/R structure and complete equality hierarchy
\end{enumerate}

The connection between truth and RG flow within the L/B/R structure and complete equality hierarchy provides a natural bridge between logic and physics. In the next section, we will develop the effective logic framework that unifies all computational paradigms through the MDE pyramid structure.
